---
title: "Assignment 01"
---
1. Review/refresh R programming:

  * Resources:
    1. [Introduction to R (University of Southern California)](https://www.uscspec.org/introduction-to-r)
    <p style="color:green">Completed, no exercises.</p>
    
    2. [EPPS Math and Coding Camp (Start from Module 3)](https://eppsmathcodingcamp.github.io/modules/)
    <p style="color:green">See below for exercises.</p>
    
2. Project
    <p style="color:green">Please refer to [Project Brainstorming](https://jmglendenning.github.io/project01.html "EPPS 6323 Project") for information on the project for EPPS 6323.  
    </p>
  * Brainstorm:
    1. Propose a hypothesis
    2. How data can be collected?
    3. What methods could be considered?
    4. How to start the data project

# Exercises - Review / Refresh R Skills

These exercises are from Day 3 & 4 of [EPPS Math Coding Camp](https://eppsmathcodingcamp.github.io/modules/ "Summer Math and Coding Camp")

A few lines of code were generated with an LLM. For instance, I had forgotten how to 'paste' in R. They were also used to generate names and for help functions.

# Getting Started with R and RStudio

## Exercise 1

After installation, type R.version. What version of R did you install? What is the nickname of that particular software build?

```{r}

R.version

```

## Exercise 2

Open RStudio. In the console pane (usually at the bottom-left), type 1 + 1 and press Enter. What result do you get?

```{r}

1+1

```

## Exercise 3

Create a new R script (File \> New File \> R Script). Type print("Hello, Data Science!") and run the code. What output do you see in the console?

```{r}

print("Hello, Data Science!")

```

## Exercise 4

Use pacman to install and load the tidyr package. Then, use p_functions() to list all functions in the tidyr package.

```{r}
library(pacman)
p_functions(tidyr)
```

## Exercise 5

Create a new folder on your computer called “DataScience”. Set this as your working directory in RStudio. Then, use getwd() to confirm it’s set correctly.

Note: This was done, but running this in Quarto is problematic, so this output is copied from the Console.

```         
> getwd()
[1] "/Users/john/DataScience"
```

